### 2021-06-03 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Edits to fig layouts mainly
Further edits. New spy plots to make uniform across thesis. New references and fixing of capitals etc.

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the Spherical Harmonics as polynomials in three variables $x, y, z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why this is a useful way to define our domain.

In this chapter, we demonstrate how one can write the Spherical Harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the Spherical Harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere.

@@

Jacobi operators are matrices that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x, y, z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere. Relations concerning the products of Spherical Harmonics have already been established, and so the expressions we desire here could just be seen as special cases of these. More specifically, the coefficients of the expansion in the SH basis of the product of two SHs are composed of Clebsch-Gordan coefficients \cite[p231]{sakurai1995modern}. Clebsch-Gordan coefficients are important in quantum mechanics and angular momentum, with tables and algorithms existing to calculate them \cite{mcnamee1964tables}. By simply choosing appropriate combinations, one could gain the desired coefficients for the expansion of each of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$.

However we reiterate our goal here is to use techniques that can generalise to spectral methods on other domains using multidimensional OPs, and thus we proceed by exploiting the three-term recurrences and other relations of the univariate OPs that make up the Spherical Harmonics as we have defined them. With that being said, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$ the following identities hold:
\bseqn{
	&1)\quad P_{n}^{(\alpha, \beta)}(z) = {1 \over 2n + \alpha + \beta + 1} \Big\{- {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  					& \quadfour \quadtwo \quad + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									 					& \quadfour \quadtwo \quad + {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\ \\
	&2)\quad (1-z^2) \, P_{n}^{(\alpha, \beta)}(z) = {4 \over 2n + \alpha + \beta + 1} \Big\{{(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\ \\
	&3)\quad z \, P_{n}^{(\alpha, \beta)}(z) = {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
								& \quadeight + {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
								& \quadeight + {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
}
\end{lemma}
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a useful corollary specific for our needs.
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
\bseqn{
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
where
\bseqn{
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1}, \quad
	&&\tilde \alpha_{l,m,2} := 
		\begin{cases}
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
		\end{cases}, \\
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)}, \quad
	&&\tilde \alpha_{l,m,4} := {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)}, \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases}, \quad
	&&\tilde \gamma_{l,m,2} := \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
}
\end{corollary}
\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

@@

\remark The results in \bsreflemma{lemma:shxyzrelations} are fairly involved and can be derived via the relationships established for products of Spherical Harmonics and Clebsch-Gordan coefficients as mentioned earlier \cite[p231]{sakurai1995modern} but we include this proof so as to generalise to other OPs later on in later chapters.

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jx}
		\centering
		\caption{$J_x$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jy}
		\centering
		\caption{$J_y$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jz}
		\centering
		\caption{$J_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices, showing their sparsity and structure, up to order $N=10$.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

Three-term recurrence relations for orthogonal polynomials are well established (see for example \cite[Section 18.9]{DLMF}), including for multidimensional OPs \cite[Section 3.2]{dunkl2014orthogonal}. In a similar vein, we can obtain a recurrence relation for the Spherical Harmonics by combining each system in \bsrefeqn{eqn:jacobimatsdef} in a specific way. It should be emphasised at this point that this is not necessarily the optimal way of computing the Spherical Harmonics; various efficient methods and algorithms have been proposed (see e.g. \cite{guy1998efficient, sloan2013efficient, fukushima2012numerical}). We present this method in order to be able to generalise it to other multidimensional OP families later on in this thesis.

@@

We have established that the Spherical Harmonics can be viewed as a basis of orthonormal polynomials on the sphere. Specifically, given a function $f : \Omega \to \C$, we can write $f = \sum_{l = 0}^{N} \sum_{m = -l}^{l} f_{l,m} \; \Ylm$ where $\{f_{l,m}\}$ are coefficients and $N \in \N$ is sufficiently large. But what about vector-valued functions?
\begin{definition}\label{def:tangentbundle}
	Let $(x,y,z) \in \Omega$. We refer to the \textbf{tangent space} of the point $(x,y,z)$ as the space of all vectors that are orthogonal to the vector $\rvec := [x,y,x]^\top$. We further define the \textbf{tangent bundle} of $\Omega$ as the set of all such spaces for every point in $\Omega$.
\end{definition}
For the purposes of this discussion, we will only consider the tangent bundle, and thus wish to write down a basis for vector-valued functions with values in the tangent bundle. Fortunately, there is a simple solution known as the \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}:
\bseqnnumber{
	\gradYlm &:= \nabla \Ylm = \ppx{\phi} \Ylm \phivec + {1 \over \sinphi} \ppx{\theta} \Ylm \thetavec, \nonumber \\
	\gradpYlm &:= \nabla^\perp \Ylm \equiv \rvec \times \nabla \Ylm = \ppx{\phi} \Ylm \thetavec - {1 \over \sinphi} \ppx{\theta} \Ylm \phivec. \label{eqn:VSHs}
}
Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$, while $\phivec, \thetavec$ are the standard spherical coordinate unit vectors, i.e. $\phivec := [\costheta \cosphi, \sintheta \cosphi, -\sinphi]^\top$, $\thetavec := [-\sintheta, \costheta, 0]^\top$. They are orthogonal with respect to the inner product $\ip<\vec{A}, \vec{B}> = \iint_\Omega \vec{A} \cdot \vec{B}^* \: \sinphi \: \D\varphi \: \D\theta$ -- in particular, for any $l, l' \in \N$, $m, m' \in \Z$ s.t. $-l \le m \le l$, $-l' \le m' \le l'$,
\bseqn{
	\ip<\gradYlm, \gradYlmprime> &= \iint_\Omega \gradYlm \cdot {\gradYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradpYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = 0.
} 
Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent bundle} of the sphere. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for such functions \cite{barrera1985vector}. That is, given $\uvec$, there exist coefficients $\{u_{l,m}^{\gradYlm}, u_{l,m}^{\gradpYlm} \}$ such that for large enough $N$:
\bseqn{
	\uvec = \sum_{l = 0}^{N} \sum_{m = -l}^{l} u_{l,m}^{\gradYlm} \; \gradYlm + u_{l,m}^{\gradpYlm} \; \gradpYlm.
}
Vector spherical harmonics have been widely used in electrostatics (e.g. \cite{barrera1985vector}), electrodynamics (e.g. \cite{carrascal1991vector}), and of course fluid dynamics including weather and climate modelling (e.g. \cite{moses1974use, fengler2005nonlinear, swarztrauber1996spectral}). Other definitions for VSHs are used (e.g. \cite{hill1954theory}), however it is convenient for deriving explicit sparse relations and operators to use the ones described here. In particular, we can note that computing the gradient or divergence of a function is merely a matter of conversion between the SH and the VSH bases.

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jx-tangent}
		\centering
		\caption{$\tangentjacobi_x$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jy-tangent}
		\centering
		\caption{$\tangentjacobi_y$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jz-tangent}
		\centering
		\caption{$\tangentjacobi_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices for the Vector Spherical Harmonic OP basis, showing their sparsity and structure, up to order $N=10$.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

\begin{figure}[tp]
	\centering % <-- added
	\centerline{\includegraphics[scale=0.6]{heat-eqn-alpha-0p0238-dt=0p01-nsteps=100}}
	\caption{Four snapshots of the solution $u$ of the heat equation with $\alpha = 1/42$ and a step size of $\Delta t = 0.01$, solved using the BDF2 timestepping method. Top Left: Initial conditions (0 timesteps). Top Right: After 25 timesteps. Bottom Left: After 50 timesteps. Bottom Right: After 75 timesteps.}
	\label{fig:heateqn}
\end{figure}

@@

\begin{figure}[tp]
	\centering % <-- added
	\centerline{\includegraphics[scale=0.6]{linear-swe-H=1-dt=0p01-nsteps=100}}
	\caption{Four snapshots of the solution for the height $h$ of the linearised shallow water equations with $\rotrate = \calH = 1$ and a step size of $\Delta t = 0.01$, solved using the backwards Euler timestepping method. Top Left: Initial conditions (0 timesteps). Top Right: After 50 timesteps. Bottom Left: After 75 timesteps. Bottom Right: After 100 timesteps.}
	\label{fig:swe}
\end{figure}

@@

As a demonstration, in \bsreffig{fig:swe} we see four snapshots of the solution to the system in \bsrefeqn{eqn:swe} on the unit sphere, with $\rotrate = \calH = 1$ and initial conditions given by
\bseqn{
	\uvec_0 := \phivec \mu \sinphi, \quad h_0 := \mu \sin^2\varphi,
}
where $\mu = {\pi \over 6}$, with the solution order $N$ set large enough (here, $N=20$). This example is very similar to \enquote{test case 2} in the well-known standard test set for shallow water equations in spherical geometries \cite{williamson1992standard}. The snapshots of the wave height $h$ are taken of the initial conditions, plus after $50$, $75$ and $100$ timesteps respectively, using a step size of $\Delta t = 0.01$.

>>>>

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x, y, z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why this is a useful way to define our domain.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere.

@@

Jacobi operators are matrices that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x, y, z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere. Relations concerning the products of spherical harmonics have already been established, and so the expressions we desire here could just be seen as special cases of these. More specifically, the coefficients of the expansion in the SH basis of the product of two SHs are composed of Clebsch-Gordan coefficients \cite[p231]{sakurai1995modern}. Clebsch-Gordan coefficients are important in quantum mechanics and angular momentum, with tables and algorithms existing to calculate them \cite{mcnamee1964tables}. By simply choosing appropriate combinations, one could gain the desired coefficients for the expansion of each of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$.

However we reiterate our goal here is to use techniques that can generalise to spectral methods on other domains using multidimensional OPs, and thus we proceed by exploiting the three-term recurrences and other relations of the univariate OPs that make up the spherical harmonics as we have defined them. With that being said, we require a few results concerning the incrementing and decrementing of the parameters for the Jacobi polynomials, and the three-term recurrence for the Jacobi polynomials, that will also prove useful later on in this chapter.
\begin{lemma}
	Let $\alpha, \beta \in \R$ be general parameters. Let $z \in [-1,1]$ and define $P_{-2}^{(\alpha+1, \beta+1)}(z) \equiv P_{-1}^{(\alpha+1, \beta+1)}(z) \equiv 0$. Then, for any $n \in \N_0$ the following identities hold:
\bseqn{
	&1)\quad P_{n}^{(\alpha, \beta)}(z) = {1 \over 2n + \alpha + \beta + 1} \Big\{- {(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n-2}^{(\alpha+1, \beta+1)}(z) \\
									  					& \quadfour \quadtwo \quad + (n+\alpha+\beta+1) \Big[{n+\alpha \over 2n+\alpha+\beta} - {n+\beta+1 \over 2n+\alpha+\beta+2}\Big] \: P_{n-1}^{(\alpha+1, \beta+1)}(z) \\
									 					& \quadfour \quadtwo \quad + {(n+\alpha+\beta+1)(n+\alpha+\beta+2) \over 2n+\alpha+\beta+2} \: P_{n}^{(\alpha+1, \beta+1)}(z) \Big\}, \\ \\
	&2)\quad (1-z^2) \, P_{n}^{(\alpha, \beta)}(z) = {4 \over 2n + \alpha + \beta + 1} \Big\{{(n+\alpha)(n+\beta) \over 2n+\alpha+\beta} \: P_{n}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: + (n+1) \Big[{n+\alpha+1 \over 2n+\alpha+\beta+2} - {n+\beta \over 2n+\alpha+\beta}\Big] \: P_{n+1}^{(\alpha-1, \beta-1)}(z) \\
									 							& \quadeight \quadtwo \: - {(n+1)(n+2) \over 2n+\alpha+\beta+2} \: P_{n+2}^{(\alpha-1, \beta-1)}(z) \Big\}, \\ \\
	&3)\quad z \, P_{n}^{(\alpha, \beta)}(z) = {2(n+\alpha)(n+\beta) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 1)} \: P_{n-1}^{(\alpha, \beta)}(z) \\
								& \quadeight + {(\beta^2 - \alpha^2) \over (2n + \alpha + \beta)(2n + \alpha + \beta + 2)} \: P_{n}^{(\alpha, \beta)}(z) \\
								& \quadeight + {2(n+1)(n + \alpha + \beta + 1) \over 2(2n + \alpha + \beta + 1)(2n + \alpha + \beta + 2)} \: P_{n+1}^{(\alpha, \beta)}(z).
}
\end{lemma}
\begin{proof}
	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}
\end{proof}
We thus can present a useful corollary specific for our needs.
\begin{corollary}\label{corollary:jacobirelations}
	Let $l, m \in \N_0$ s.t. $0 \le m \le l$ and let $z \in [-1,1]$. Then:
\bseqn{
	(1-z^2) \, \jaclm{l}{m}(z) &=  \tilde \alpha_{l,m,1} P^{(m-1,m-1)}_{l-m}(z) + \tilde \alpha_{l,m,3} P^{(m-1,m-1)}_{l-m+2}(z) \\
	\jaclm{l}{m}(z) &= \tilde \alpha_{l,m,2} P^{(m+1,m+1)}_{l-m-2}(z) + \tilde \alpha_{l,m,4} P^{(m+1,m+1)}_{l-m}(z), \\
	z \, \jaclm{l}{m}(z) &= \tilde \gamma_{l,m,1} P^{(m,m)}_{l-m-1}(z) + \tilde \gamma_{l,m,2} P^{(m,m)}_{l-m+1}(z),
}
where
\bseqn{
	\tilde \alpha_{l,m,1} &:= {2l \over 2l+1}, \quad
	&&\tilde \alpha_{l,m,2} := 
		\begin{cases}
			- {l \over 2(2l+1)} \quad \text{if } l - m \ge 2 \\
			\quad \quad 0 \quad \quad \text{otherwise}
		\end{cases}, \\
	\tilde \alpha_{l,m,3} &:= - {2(l-m+2)(l-m+1) \over (2l+1)(l+1)}, \quad
	&&\tilde \alpha_{l,m,4} := {(l+m+2)(l+m+1) \over 2(2l+1)(l+1)}, \\
	\tilde \gamma_{l,m,1} &:= \begin{cases}
			\frac{l}{2l+1} \quad \text{if } l - m \ge 1 \\
			\quad 0 \quad \quad \text{otherwise}
		\end{cases}, \quad
	&&\tilde \gamma_{l,m,2} := \frac{(l-m+1)(l+m+1)}{(2l+1)(l+1)}.
}
\end{corollary}
\bsrefcor{corollary:jacobirelations} allows us to easily determine the aforementioned coefficients, that we will formalise in the following Lemma.

@@

\remark The results in \bsreflemma{lemma:shxyzrelations} are fairly involved and can be derived via the relationships established for products of spherical harmonics and Clebsch-Gordan coefficients as mentioned earlier \cite[p231]{sakurai1995modern} but we include this proof so as to generalise to other OPs later on in later chapters.

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jx}}
		\centering
		\caption{$J_x$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jy}}
		\centering
		\caption{$J_y$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jz}}
		\centering
		\caption{$J_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices, showing their sparsity and structure, up to order $N=10$.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

Three-term recurrence relations for orthogonal polynomials are well established (see for example \cite[Section 18.9]{DLMF}), including for multidimensional OPs \cite[Section 3.2]{dunkl2014orthogonal}. In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in \bsrefeqn{eqn:jacobimatsdef} in a specific way. It should be emphasised at this point that this is not necessarily the optimal way of computing the spherical harmonics; various efficient methods and algorithms have been proposed (see e.g. \cite{guy1998efficient, sloan2013efficient, fukushima2012numerical}). We present this method in order to be able to generalise it to other multidimensional OP families later on in this thesis.

@@

We have established that the spherical harmonics can be viewed as a basis of orthonormal polynomials on the sphere. Specifically, given a function $f : \Omega \to \C$, we can write $f = \sum_{l = 0}^{N} \sum_{m = -l}^{l} f_{l,m} \; \Ylm$ where $\{f_{l,m}\}$ are coefficients and $N \in \N$ is sufficiently large. But what about vector-valued functions?
\begin{definition}\label{def:tangentbundle}
	Let $(x,y,z) \in \Omega$. We refer to the \textbf{tangent space} of the point $(x,y,z)$ as the space of all vectors that are orthogonal to the vector $\rvec := [x,y,x]^\top$. We further define the \textbf{tangent bundle} of $\Omega$ as the set of all such spaces for every point in $\Omega$.
\end{definition}
For the purposes of this discussion, we will only consider the tangent bundle, and thus wish to write down a basis for vector-valued functions with values in the tangent bundle. Fortunately, there is a simple solution known as the \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}:
\bseqnnumber{
	\gradYlm &:= \nabla \Ylm = \ppx{\phi} \Ylm \phivec + {1 \over \sinphi} \ppx{\theta} \Ylm \thetavec, \nonumber \\
	\gradpYlm &:= \nabla^\perp \Ylm \equiv \rvec \times \nabla \Ylm = \ppx{\phi} \Ylm \thetavec - {1 \over \sinphi} \ppx{\theta} \Ylm \phivec. \label{eqn:VSHs}
}
Here, $\rvec$ is simply the outward unit normal vector to the surface of the sphere at the point $(x,y,z)$, while $\phivec, \thetavec$ are the standard spherical coordinate unit vectors, i.e. $\phivec := [\costheta \cosphi, \sintheta \cosphi, -\sinphi]^\top$, $\thetavec := [-\sintheta, \costheta, 0]^\top$. They are orthogonal with respect to the inner product $\ip<\vec{A}, \vec{B}> = \iint_\Omega \vec{A} \cdot \vec{B}^* \: \sinphi \: \D\varphi \: \D\theta$ -- in particular, for any $l, l' \in \N$, $m, m' \in \Z$ s.t. $-l \le m \le l$, $-l' \le m' \le l'$,
\bseqn{
	\ip<\gradYlm, \gradYlmprime> &= \iint_\Omega \gradYlm \cdot {\gradYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradpYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = \delta_{l,l'} \delta_{m,m'} \: l (l+1), \\
	\ip<\gradYlm, \gradpYlmprime> &= \iint_\Omega \gradpYlm \cdot {\gradpYlmprime}^* \: \sinphi \: \D\varphi \: \D\theta = 0.
} 
Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent bundle} of the sphere. Importantly, the set $\{\gradYlm, \gradpYlm\}$ form a complete and orthogonal basis for such functions \cite{barrera1985vector}. That is, given $\uvec$, there exist coefficients $\{u_{l,m}^{\gradYlm}, u_{l,m}^{\gradpYlm} \}$ such that for large enough $N$:
\bseqn{
	\uvec = \sum_{l = 0}^{N} \sum_{m = -l}^{l} u_{l,m}^{\gradYlm} \; \gradYlm + u_{l,m}^{\gradpYlm} \; \gradpYlm.
}
Vector spherical harmonics have been widely used in electrostatics (e.g. \cite{barrera1985vector}), electrodynamics (e.g. \cite{carrascal1991vector}), and of course fluid dynamics including weather and climate modelling (e.g. \cite{moses1974use, fengler2005nonlinear, swarztrauber1996spectral}). Other definitions for VSHs are used (e.g. \cite{hill1954theory}), however it is convenient for deriving explicit sparse relations and operators to use the ones described here. In particular, we can note that computing the gradient or divergence of a function is merely a matter of conversion between the SH and the VSH bases.

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jx-tangent}}
		\centering
		\caption{$\tangentjacobi_x$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jy-tangent}}
		\centering
		\caption{$\tangentjacobi_y$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jz-tangent}}
		\centering
		\caption{$\tangentjacobi_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices for the vector spherical harmonic OP basis, showing their sparsity and structure, up to order $N=10$.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

\begin{figure}[t]
	\centering % <-- added
	\centerline{\includegraphics[scale=0.5]{heat-eqn-alpha-0p0238-dt=0p01-nsteps=100}}
	\caption{Four snapshots of the solution $u$ of the heat equation with $\alpha = 1/42$ and a step size of $\Delta t = 0.01$, solved using the BDF2 timestepping method. Top Left: Initial conditions (0 timesteps). Top Right: After 25 timesteps. Bottom Left: After 50 timesteps. Bottom Right: After 75 timesteps.}
	\label{fig:heateqn}
\end{figure}

@@

\begin{figure}[t]
	\centering % <-- added
	\centerline{\includegraphics[scale=0.5]{linear-swe-H=1-dt=0p01-nsteps=100-arrows}}
	\caption{Four snapshots of the solution for the height $h$ (surface colour) and the wave velocity $u$ (red arrows) of the linearised shallow water equations with $\rotrate = \calH = 1$ and a step size of $\Delta t = 0.01$, solved using the backwards Euler timestepping method (BDF1). The initial conditions are given in \bsrefeqn{eqn:sweics}. Top Left: Initial conditions (0 timesteps). Top Right: After 50 timesteps. Bottom Left: After 75 timesteps. Bottom Right: After 100 timesteps.}
	\label{fig:swe}
\end{figure}

@@

As a demonstration, in \bsreffig{fig:swe} we see four snapshots of the solution to the system in \bsrefeqn{eqn:swe} on the unit sphere, with $\rotrate = \calH = 1$ and initial conditions given by
\bseqnnumber{
	\uvec_0 := \phivec \mu \sinphi, \quad h_0 := \mu \sin^2\varphi, \label{eqn:sweics}
}
where $\mu = {\pi \over 6}$, with the solution order $N$ set large enough (here, $N=20$). This example is very similar to \enquote{test case 2} in the well-known standard test set for shallow water equations in spherical geometries \cite{williamson1992standard}. The snapshots of the wave height $h$ are taken of the initial conditions, plus after $50$, $75$ and $100$ timesteps respectively, using a step size of $\Delta t = 0.01$.
